Live Functional Programming with Typed Holes

### 2018-04-16 calculus.tex
intro edits

\subsubsection{Complete Programs} 
%
Although this paper focuses running \emph{incomplete} programs, it helps
to know that its machinery does not interfere
with reasoning about the behavior of \emph{complete} programs, i.e. those
with no holes.
%
Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
%
Failed casts cannot appear in complete internal expressions. 
%
The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

>>>>

\subsubsection{Complete Programs} 
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that its machinery does not interfere
with reasoning about the behavior of \emph{complete} programs, i.e. those
with no holes.
%
Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
%
Failed casts cannot appear in complete internal expressions. 
%
The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

### 2018-04-16 intro.tex
intro edits

% \matt{ ``The problem that specifically motivates this paper'' is a
% really long noun phrase; the sentence is more interesting, IMO, when
% the subject is 'programming languages' and not that noun phrase.  }
%
%The problem that specifically motivates this paper \IS that
Problematically,
programming languages typically assign dynamic meaning only to {complete programs}, i.e. programs that are syntactically well-formed and free of static type and binding errors. A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states. As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e. it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer
%
%\IS actually seeing and writing in the editor.
actually sees and writes in the editor.
\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}

In some cases these gaps are momentary, like while the programmer
%\IS entering
enters
a short expression. In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, when the programmer changes a type definition, or when the programmer makes a mistake.
%
Further, such mistakes arise more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}.
%
%\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
%The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program. Throughout the process of addressing these errors, the program text remains formally meaningless. 
Overall, an analysis of edits performed by Java programmers using Eclipse found that about 40\% of edit states were malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.

@@

%\matt{ Below, notice the broadening from ``the problem that motivates
%this paper is that incomplete lack dynamic meaing'' to just the
%general quesiton of ``incomplete program meaning''--- another reason
%to do my edit above, and remove that specific phrase.
%}
In recognition of this problem---that incomplete programs lack meaning---%
\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
functional programs, modeled formally as typed expressions with \emph{holes} in 
both expression and type position. 
Empty holes stand for missing expressions or types,
and non-empty holes operate as ``membranes'' around static type inconsistencies 
(i.e. they internalize the ``red underline'' that editors commonly display under a type inconsistency).
% For editor states into this language of incomplete programs, 
% editor services can reason about types and binding in many more situations than previously possible.
\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from an editor state. 
Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
the programmer can insert explicitly holes manually (see below) 
or semi-automatically via code completion techniques \cite{Amorim2016}, 
or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).
%
For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes.
%
%Our goal in this paper \IS to develop

%%%%%%%%%%%% New par:
To address this need,
this paper develops
a \emph{dynamic semantics} for incomplete functional programs.
%
To understand this problem, consider a (common) approach in practice, where evaluation simply stops with an error when it encounters an ``expression hole,'' encoded by the programmer with a raised exception.
%This mirrors a workaround that programmers commonly deploy: 
%raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
Some languages provide more concise notation for this interpretation of expression holes, 
e.g. Scala uses \lismall{???} for missing method bodies 
and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u}
%where \lismall{u} \IS an
where \lismall{u} serves as an
optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
\footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
%This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave this way.}

Although better than nothing, this approach of stopping with an exception upon reaching an expression hole has limitations 
as the basis for a live programming environment because 
(1)~it provides limited information about the dynamic state of the program when the exception occurs
(typically only a stack trace);  
(2)~it provides no information about the behavior of the remainder of program, 
parts of which may not depend at all on the missing or erroneous expression; and 
(3)~it provides no means by which to resume evaluation after filling the hole.

Moreover, we need to be able to run programs containing not only expression holes but also type holes.
The static semantics developed by \citet{popl-paper} derives the machinery for 
reasoning statically about type holes from gradual type theory, 
identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}.
%
As such, we can look to the dynamic semantics from
gradual type theory to specify the dynamic semantics for incomplete programs,
which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
%However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
However, when a cast fails, evaluation again stops with 
an exception.
%
As a result, traditional gradual typing approaches still leave the live programming environment unable to provide feedback about the parts 
of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).

@@

We are integrating this approach into the \Hazel programming environment being developed independently of this paper by \citet{HazelnutSNAPL}. 
We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several examples, in the setting of this specific design.

Sec.~\ref{sec:calculus} then abstracts away orthogonal details and makes these intuitions formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions as just outlined, in a manner that admits clean type safety theorems (in particular, a Progress theorem). The semantics of \HazelnutLive borrows machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}
(to handle incomplete types) and contextual modal type theory~\cite{Nanevski2008} (CMTT,
which provides a logical foundation for hole environment tracking). 
These non-obvious connections to well-established existing systems help support our claim that this approach \IS theoretically well-grounded\todo{note that our exposition \IS self-contained here? or later?}.

Sec.~\ref{sec:agda-mechanization} outlines our mechanization of the core calculus using the Agda proof assistant. 
Sec.~\ref{sec:implementation} outlines our simple implementation of the core calculus, which implements the live programming features from Sec.~\ref{sec:examples},
but for the more austere language of Sec.~\ref{sec:calculus}. 
The editor component of this implementation provides a language of structured edit actions, 
based on the \Hazelnut structure editor calculus developed by \cite{popl-paper}, that guarantees that
every editor state has some, possibly incomplete, type. 
Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result \IS a proof-of-concept
live functional programming environment where every possible editor state has both non-trivial static and dynamic 
meaning, i.e. live feedback never ``flickers'' or ``goes out''. 
The mechanization and the implementation are both available in the supplemental material.

Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which \IS rooted in the contextual substitution operation from CMTT. We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation for cells in a live lab notebook environment.

Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneous discusses directions for future work. Sec.~\ref{sec:discussion} briefly concludes. The appendix includes some straightforward auxiliary definitions and proofs mentioned in the paper, some extensions to the core calculus, and some screenshots and additional details on the implementation.

>>>>

% \matt{ ``The problem that specifically motivates this paper'' is a
% really long noun phrase; the sentence is more interesting, IMO, when
% the subject is 'programming languages' and not that noun phrase.  }
%
The problem at the heart of this paper is that
programming languages typically assign meaning only to {complete programs}, i.e. programs that are syntactically well-formed and free of static type and binding errors. A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states. As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e. it relies on the most recent complete editor state, creating a perceptive gap because the feedback may not accurately reflect what the programmer is actually seeing and writing in the editor.
\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}

In some cases these gaps are momentary, like while the programmer
%\IS entering
enters
a short expression. In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank or when the programmer makes a mistake.
%
Novice programmers, of course, make more mistakes \cite{mccauley2008debugging,fitzgerald2008debugging}.
%
%\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
The problem is particularly pronounced for languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program. Throughout the process of addressing these errors, the program text remains formally meaningless. 
Overall, an analysis of edits performed by Java programmers using Eclipse found that about 40\% of edit states were malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.

@@

%\matt{ Below, notice the broadening from ``the problem that motivates
%this paper is that incomplete lack dynamic meaing'' to just the
%general quesiton of ``incomplete program meaning''--- another reason
%to do my edit above, and remove that specific phrase.
%}
In recognition of this problem---that incomplete programs lack meaning---%
\citet{popl-paper} develop a static semantics (i.e. a type system) for incomplete 
functional programs, modeling them formally as typed expressions with \emph{holes} in 
both expression and type position. 
Empty holes stand for missing expressions or types,
and non-empty holes operate as ``membranes'' around static type inconsistencies 
(i.e. they internalize the ``red underline'' that editors commonly display under a type inconsistency).
% For editor states into this language of incomplete programs, 
% editor services can reason about types and binding in many more situations than previously possible.
\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from an editor state. 
Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
the programmer can insert explicit holes manually (see below) 
or semi-automatically via code completion techniques \cite{Amorim2016}, 
or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).
%
For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes. This paper aims to address this need by developing
a {dynamic semantics} for incomplete functional programs, 
starting from the static semantics of \citet{popl-paper}.
%
%Our goal in this paper \IS to develop

One approach would be to define a dynamic semantics that simply aborts with an error when evaluation reaches a hole. 
%%%%%%%%%%%% New par:
%
This mirrors a workaround to the problem that programmers commonly deploy: 
raising an exception where an expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
Some languages provide more concise notation for this ``encoding'' of expression holes, 
e.g. Scala provides \lismall{???}  
and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u}
where \lismall{u} serves as an
optional hole name \cite{GHCHoles}.%
\footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
Proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} also respond to holes in this way.}

Although better than nothing, this approach of stopping with an exception upon reaching an expression hole has limitations 
as the basis for a live programming environment because 
(1)~it provides limited information about the dynamic state of the program when the exception occurs
(typically only a stack trace);  
(2)~it provides no information about the behavior of the remainder of program, 
parts of which may not depend on the missing or erroneous expression (e.g. later cells in a lab notebook); and 
(3)~it provides no means by which to resume evaluation after filling a hole.

Moreover, we need to be able to run programs containing not only expression holes but also type holes.
The static semantics developed by \citet{popl-paper} derives the machinery for 
reasoning statically about type holes from gradual type theory, 
identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}.
%
As such, we can look to the dynamic semantics from
gradual type theory to specify the dynamic semantics for incomplete programs,
which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
%However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
However, when a cast fails, evaluation again stops with 
an exception.
%
As a result, traditional gradual typing approaches still leave the live programming environment unable to provide feedback about the parts 
of the program that do not depend on the problematic sub-expression.

@@

We are integrating this approach into the \Hazel programming environment being developed (separately from this paper) by \citet{HazelnutSNAPL}. 
We begin in Sec.~\ref{sec:examples} by detailing the approach informally, with several examples, in the setting of this specific design.

Sec.~\ref{sec:calculus} then abstracts away orthogonal details of the language and user interface and makes these intuitions formally precise by detailing the primary contribution of this paper: a core calculus, \HazelnutLive, that supports evaluating incomplete expressions as just outlined, in a manner that admits clean type safety theorems (in particular, a Progress theorem). The semantics of \HazelnutLive borrows machinery from gradual type theory~\cite{DBLP:conf/snapl/SiekVCB15}
(to handle incomplete types) and contextual modal type theory~\cite{Nanevski2008} (CMTT,
which provides a logical foundation for hole environment tracking). 
These non-obvious connections to well-established existing systems help support our claim that this approach is theoretically well-grounded\todo{note that our exposition \IS self-contained here? or later?}.

Sec.~\ref{sec:agda-mechanization} outlines our mechanization of the core calculus using the Agda proof assistant. 
Sec.~\ref{sec:implementation} outlines our simple implementation of the core calculus, which implements the live programming features from Sec.~\ref{sec:examples},
but for the more austere language of Sec.~\ref{sec:calculus}. 
The editor component of this implementation provides a language of structured edit actions, 
based on the \Hazelnut structure editor calculus developed by \cite{popl-paper}, that guarantees that
every editor state has some, possibly incomplete, type. 
Together with the theorems established earlier in Sec.~\ref{sec:calculus}, the result is a proof-of-concept
live functional programming environment where every possible editor state has both non-trivial static and dynamic 
meaning, i.e. live feedback never ``flickers'' or ``goes out''. 
The supplemental material includes both the mechanization and the implementation.

Sec.~\ref{sec:resumption} defines the fill-and-resume feature, which is rooted in the contextual substitution operation from CMTT. We establish the correctness of fill-and-resume with a novel commutativity property and discuss how it provides a semantic interpretation of cells in a live lab notebook environment.

Sec.~\ref{sec:relatedWork} describes related work in more detail and simultaneously discusses directions for future work. Sec.~\ref{sec:discussion} briefly concludes. The appendix includes some straightforward auxiliary definitions and proofs mentioned in the paper, some extensions to the core calculus, and some screenshots and additional details on the implementation.
